---
title: "ggplot2-intro"
author: "Paul M. Magwene"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Data frame example

```{r,echo=FALSE,warning=FALSE}
library(tidyverse)
library(viridis)
```


```{r, echo=TRUE}
head(iris)
```

## Data + Geometric Representation + Aesthetic qualities

```
library(ggplot2) # import ggplot2 functions

ggplot(data = <DATA>) + <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

```

## Example 1

```{r, echo=TRUE, fig.width=6, fig.height=4}
ggplot(data = iris) + geom_histogram(mapping = aes(x = Sepal.Length))
```

## Example 2

```{r, echo=TRUE, fig.width=6, fig.height=4}
ggplot(data = iris) + geom_boxplot(mapping = aes(x = Species,
                                                 y = Sepal.Length))
```


## Example 3

```{r, echo=TRUE, fig.width=5, fig.height=3}
ggplot(data = iris) + geom_boxplot(mapping = aes(x = Species,
                                                 y = Sepal.Length)) +
                      geom_jitter(mapping = aes(x = Species,
                                                y = Sepal.Length,
                                                color = Species),
                                  width = 0.2)
                                  
```


## Faceting (conditioning)

```{r, echo=TRUE, fig.width=3, fig.height=3}
ggplot(data = iris, aes(x = Sepal.Length, fill=Species)) + 
  geom_histogram(bins=15) + facet_wrap(~Species,ncol=1)
```


## Tour of useful `geoms`: Density plots

```{r, echo=TRUE, fig.width=6, fig.height=4}
ggplot(data = iris, aes(x = Sepal.Length, fill=Species)) + 
  geom_density(alpha=0.5)
```

## Tour of useful `geoms`: Violin plots

```{r, echo=TRUE, fig.width=5, fig.height=3}
ggplot(data = iris, aes(x = Species, 
                        y = Sepal.Length, 
                        fill=Species)) + 
  geom_violin(alpha=0.5)
```

## Tour of useful `geoms`: Scatter plots

```{r, echo=TRUE, fig.width=5, fig.height=3}
ggplot(data = iris, aes(x = Sepal.Length, 
                        y = Sepal.Width, 
                        color=Species)) + 
  geom_point(alpha=0.5)
```


## Tour of useful `geoms`: 2D density plots

```{r, echo=TRUE, fig.width=6, fig.height=2}
ggplot(data = iris, aes(x = Sepal.Length, 
                        y = Sepal.Width, 
                        color=Species)) + 
  geom_density_2d() + 
  geom_point(alpha=0.5,size=1) +
  facet_wrap(~Species)
  
```


## Tour of useful `geoms`: Line plot

```{r,echo=FALSE}
library(dplyr)
library(tidyr)
covid <- read_csv("~/Desktop/us.csv")
glimpse(covid)
```
## Tour of useful `geoms`: Line plot

```{r, echo=TRUE, fig.width=6, fig.height=2}
ggplot(data = covid, aes(x = date, y = deaths)) + 
  geom_line()
```

## Tour of useful `geoms`: Area plot

```{r, echo=TRUE, fig.width=6, fig.height=2}
ggplot(data = covid, aes(x = date, y = deaths)) + 
  geom_area(fill="firebrick")
```


## Tour of useful `geoms`: Heat maps

```{r}
iris_cor <-
  iris %>%
  select(-Species) %>%
  cor() %>%
  as_tibble(rownames="Variable_1") %>%
  pivot_longer(2:5, names_to="Variable_2",values_to = "Correlation") 

ggplot(iris_cor, 
       aes(x=Variable_1, y=Variable_2, fill=Correlation)) + 
  geom_tile()
```
```{r,fig.height=5, fig.width=8,echo=TRUE}
us_states <- read_csv("~/Desktop/us-states.csv")

SE_states <- c("North Carolina","South Carolina",
               "Arkansas","Georgia","Tennessee",
               "Louisiana","Alabama","Florida")

us_states %>%
  filter(state %in% SE_states) %>%
  arrange(state) %>%
  ggplot(aes(x=date, y=state, fill=cases_avg_per_100k)) + 
    geom_tile() +
    scale_fill_viridis(option = "C")
```

