---
title: "Bio 724: Foundations of Scientific Computing"
author: "Paul M. Magwene"
output: beamer_presentation
fontsize: "9pt"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, comment = "", collapse=TRUE, prompt=TRUE)
options(width=60)

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```


## Goals for today's class session

* Make sure you have a working R/RStudio environment
* Make sure you know how to evaluate commands in the R terminal
* Make sure you know how to create and knit an RMarkdown document
* Review R data types -- numerical (double, integer, complex), logical, character 
* Review R data structures -- vectors, lists, data frames

## RStudio Tour

In-class demonstration of RStudio interface.


## RMarkdown 

In-class demonstration of creating and knitting an RMarkdown document.

## Data types


- Refers to the types of values that can be represented in a computer program
- Determine the representation of values in memory
- Constrains the operations you can perform on those values

In R data types are usually inferred by the interpreter rather than specified by the user, though occasionally you may find it necessary to specify a data type.

## Numeric data types in R

- Floating point values (`doubles') -- represent real numbers (continuous values)
```{r}
x <- 10.0
typeof(x) 
```

- Integers -- represent whole numbers. The default numeric type is double, so you must explicitly ask for integers.

```{r}
x <- as.integer(10)
typeof(x)
```

- Complex numbers -- numbers with a real and "imaginary" part. We won't be explicitly using complex numbers in this course, but they sometimes appear unexpectedly in some calculations.
```{r}
x <- 1+1i
typeof(x)
```

## Arithmetic operations on numerical data types

```{r,size="small"}
10 + 2 # addition
10 - 2 # subtraction
10 * 2 # multiplication
10 / 2 # division
10 ^ 2 # exponentiation
10 ** 2 # alternate exponentiation
10 %% 4 # modulus (remainder after division)
4-5 / 2 # operator precedence matters!
(4-5)/2 # parentheses help you specify/disambiguate precedence.
```

## Basic mathematical functions


```{r,size="small"}
sqrt(10) # square root
10 ^ 0.5 # same as square root
sqrt(-1) # NaN means "Not a Number"
sqrt(-1 + 0i) # But works if we use complex type
exp(1) # exponential function
log(100)  # log base e
log10(100) # log base 10
log2(8) # log base 2
factorial(5) # factorial function: 5 * 4 * 3 * 3 * 1
pi  # R knows some useful constants
cos(2*pi)  # cosine, also sin, tan, ...
```

## Logical (Boolean) type

```{r,size="small"}
x <- TRUE
typeof(x)
y <- FALSE
typeof(FALSE)
!TRUE # logical negation of TRUE
!y # logical negation (NOT) of value in y
```

Numerical comparison operators return logical types:

```{r,size="small"}
4 < 5  # less than
10 >= 9 # greater than or equal to
```


Comparison operations can be assigned to variables:

```{r,size="small"}
x <- 1 > 2
y <- 2 >= 2
x & y # Logical AND
x | y # Logical OR
isTRUE(x) # returns true if x is logical and true
```


## Character (string) type

```{r}
x <- "Hello" # enclosed in double quotes
y <- 'World' # or single quotes
z <- 'You said "Hello World"' # allows nesting
```

Simple character functions:

```{r}
paste(x, y)  # concatenate strings
paste(x, y, sep = "") # concatenate with no space
strsplit("Hello world!", split=" ") # split on space
```

The `stringr` package contains many useful string manipulation functions.


## Data structures


- Represent different ways collections of data are stored in memory or accessed by the user

- Homogenous or heterogeneous with respect to data types

- Different data structures are more efficient for particular modes of access or to represent different types operations

## Vectors

- Homogeneous ordered list of items

```{r}
x <- c(2, 4, 6, 8) 
y <- c("hello", "world", "how", "are", "you?")
```

Vectors always have a length (possibly zero) and a type

```{r}
length(x)
typeof(x)
length(y)
typeof(y)
```

## Elements of a vector will be coerced to be the same type

```{r}
x <- c(1+1i, 2+1i, 'Fred', 10)
x
typeof(x)
```


## Indexing vectors

Accessing the objects in a vector is accomplished by "indexing".

The elements of the vector are assigned indices $1 \ldots n$ where $n$ is the length of the vector

```{r}
x <- c(2,4,6,8)
length(x)
x[1]
x[2]
x[4]
x[length(x)] # why might this be preferred way to get last value?
```

Indexing past the end of a vector returns `NA` (Not Available) values:

```{r}
x[5]
```

## Single objects of core data types in R are themselves vectors

```{r}
x <- 1
length(x)
x[1] # can be indexed like any other vectors
is.vector(x)  # function to test whether something is vector
```



## Arithmetic operators and most math functions work on numerical vectors

Arithmetic operators and comparison work element-by-element on vectors.

```{r, size="small"}
x <- c(2, 4, 6, 8)
y <- c(0, 1, 2, 10)
x + y
x * y
x^2
sqrt(x)
x < y
```

## Vector Recycling

When binary operators are applied to vectors of different lengths, "recycling" occurs - the shorter vector is repeated to match the length of the longer one by recycling its elements starting from the first one:

```{r}
x <- c(1, 1, 1, 1)
y <- c(0, 1)

        # original       after recylcing    result
        # x = 1 1 1 1    x = 1 1 1 1        1 2 1 2
        # y = 0 1        y = 0 1 0 1
x + y
```



## Vectors can be indexed by other vectors, including logical vectors

```{r}
x <- c(2, 4, 6, 8)
x[c(2,4)]  # get 2nd and 4th elements of x
```

```{r}
x <- c(2, 4, 6, 8)
y <- c(0, 1, 2, 10)
x[x > y] # get elements of x where x > y
```



## Lists

Lists in R are like vectors but the elements of a list are arbitrary objects (even other lists). Lists are "heteregeneous".

```{r}
x <- list('Bob',27, 10, c(720,710))
typeof(x)
x
```

## Indexing Lists

Items in lists are accessed in a  different manner than vectors.

- Typically you use double brackets (`[[]]`)to return the element at index `i`

- Single brackets always return a list containing the element at index `i`


```{r}
x <- list('Bob', 27, 10, c(720,710))
x[1]
typeof(x[1])
x[[1]]
typeof(x[[1]])
```

## List elements can have names

```{r, size="small"}
x <- list(name='Bob',age=27, years.in.school=10)
x
```

Named list objects can be accessed via the `$` operator

```{r, size="small"}
x$years.in.school
```

The names of list elements can be accessed with the `names()` function

```{r, size="small"}
names(x)
```


## Data frames

Data frames represent data tables (data sets) 
  
  - Each column in the table has the same number of rows
  - Every item in a given column has to be of the same type.
  - Columns in a data frame must have names
  
```{r}
# constructing a data frame from scratch
name <- c("Paul","Mary","Peter")
grade <- c("C","A","B")
age <- c(40, 42, 28)

example.df <- data.frame(name = name, age = age, grade = grade)
example.df
```


## Data frames: shape and column indexing

Shape 

```{r}
nrow(example.df)  # number of rows
ncol(example.df)  # number of columns
```


Indexing columns by name or position

```{r}
example.df$age
example.df[3] 
```

Indexing with a vector to subset columns
```{r}
example.df[c("name","grade")]
```

## Data frames: indexing rows

To index one or more rows of a data frame, specifying the value(s) to index on followed by a comma:

```{r}
example.df[1,]  # first row
```

```{r}
example.df[c(1,3),] # first and third row
```

You can simultaneously index both rows and columns:

```{r}
example.df[c(1,3), c("name", "age")]
```

## Matrices

In R matrices are two-dimensional collections of elements all of which have the same mode or type (homogenous). 

```{r}
r <- matrix(1:12, nrow=3, ncol = 4)
r
```

Matrices don't by default have explicit row or column names, but they can be assigned/retrieved using the `rownames()` and `colnames()` functions respectively.

```{r}
colnames(r) <- c("N", "C", "O2", "PO4")
rownames(r) <- c("Site.A", "Site.B", "Site.C")
r
```

We will typically prefer data frames to matrices, because data frames are more flexible. But we'll return to matrices when we get into linear algebra later in the semester.
